Interacting with Modal Logics \\ in the Coq Proof Assistant

### 2015-02-25 Papers/2015/ModalLogicCoq/ModalCoq.tex
Move Coq paper

\begin{abstract}  
  This paper describes an embedding of higher-order modal logics in
  the \Coq proof assistant. {\Coq}'s capabilities are extended in a
  minimalistic manner, which is nevertheless sufficient for the
  formalization of significant, non-trivial modal logic proofs. The
  elegance, flexibility and convenience of this approach, from a user
  perspective, are illustrated here with the successful formalization
  of G\"odel's ontological argument.
\end{abstract}

@@

Another recently explored possibility is the embedding of first-order
and even higher-order modal logics into classical higher-order logics
\cite{J23,B9}, for which existing higher-order automated theorem
provers \cite{LEO-II,Satallax} exist. The embedding approach is
flexible, because various modal logics (even with multiple modalities
or varying/cumulative domain quantifiers) can be easily supported by
stating their characteristic axioms. Moreover, the approach is
relatively simple to implement, because it does not require any
modification in the source code of the higher-order prover. The prover
can be used as is, and only the input files provided to the prover
must be especially encoded. Furthermore, the efficacy and efficiency
of the embedding approach has been confirmed in philosophically
relevant benchmarks \cite{FormalTheologyRepository}. These qualities
make embedding a convenient approach for \emph{fully automated}
reasoning.

@@

\noindent 
All modal connectives are simply lifted versions of the
usual logical connectives. Notations are used to allow the modal
connectives to be used as similarly as possible as the usual
connectives. The prefix ``\texttt{m}'' is used to distinguish the
modal connectives: if $\odot$ is a connective on type \text{Prop},
$\texttt{m}\odot$ is a connective on the lifted type $\texttt{o}$ of
modal propositions.

@@

\noindent
To maximally preserve user intuition in interactive modal logic
theorem proving, the embedding via the possible world semantics should
be as transparent as possible to the user. Fortunately, the basic Coq
tactics described above automatically unfold the shallowest modal
definition in the goal.  Therefore, they can be used with modal
connectives and quantifiers just as they are used with the usual
connectives and quantifiers. The situation for the new modal
operators, on the other hand, is not as simple, unfortunately.

@@

\noindent If the new modal tactics above are regarded from a natural
deduction point of view, they give rise to the inference
rules\footnote{The ND calculus with the rules from
Figures \ref{fig:NaturalDeduction} and \ref{fig:NDK} is sound and
complete relatively to the calculus of Figure
\ref{fig:NaturalDeduction} extended with a necessitation rule and the
modal axiom K \cite{ModalND}. Starting from a sound and Henkin-complete ND calculus for classical higher-order logic (cf. Figure~\ref{fig:NaturalDeduction}), the additional
modal rules in Figure \ref{fig:NDK} make it sound and Henkin-complete
for modal higher-order logic \textbf{K}.} shown in Figure
\ref{fig:NDK}. The labels that name boxes in those inference rules are
precisely the worlds that annotate goals and hypotheses in \Coq with
the modal embedding. A hypothesis of the form \texttt{(p w)}, where
\texttt{p} is a modal proposition of type \texttt{o} and \texttt{w} is
a world of type \texttt{i} indicates that \texttt{p} is an assumption
created inside a box with name \texttt{w}.

@@

\s\s
\infer[\nec_I]{\nec A}{\omega: \fbox{\infer*{A}{}} }
\s
\infer[\nec_E]{w: \fbox{ \infer*{}{A} } }{\nec A}
%\vspace{2em}
\s
\infer[\pos_I]{\pos A}{w: \fbox{\infer*{A}{}} }
\s
\infer[\pos_E]{\omega: \fbox{ \infer*{}{A} } }{\pos A}

@@

In order to demonstrate the efficacy and convenience of the modal
embedding approach not only for proving simple lemmas and theorems,
but also for larger developments, we include here a full and detailed
formalization of G\"odel's ontological argument.

@@

The theological implications of the verified correctness of G\"odel's
proof certainly depend on a critical discussion of the underlying
concepts, definitions and axioms, which is beyond the scope of this
paper.  Clearly, the application of theorem proving technology to the
domain of theoretical philosophy can --- as already pictured by
Leibniz --- be very fruitful for both areas.

>>>>

\begin{abstract}  
  This paper describes an embedding of higher-order modal logics in
  the \Coq proof assistant. {\Coq}'s capabilities are used to implement modal logics in a
  minimalistic manner, which is nevertheless sufficient for the
  formalization of significant, non-trivial modal logic proofs. The
  elegance, flexibility and convenience of this approach, from a user
  perspective, are illustrated here with the successful formalization
  of G\"odel's ontological argument.
\end{abstract}

@@

Another recently explored possibility is the embedding of first-order
and even higher-order modal logics into classical higher-order logics
\cite{J23,B9}, for which existing higher-order automated theorem
provers \cite{LEO-II,Satallax} exist. The embedding approach is
flexible, because various modal logics (even with multiple modalities
or varying/cumulative domain quantifiers) can be easily supported by
stating their characteristic axioms. Moreover, the approach is
relatively simple to implement, because it does not require any
modification in the source code of the higher-order prover. The prover
can be used as is, and only the input files provided to the prover
must be specially encoded (using lifted versions of connectives and logical constants instead of the usual ones). Furthermore, the efficacy and efficiency
of the embedding approach has been confirmed in benchmarks stemming from philosophy \cite{FormalTheologyRepository}. These qualities
make embedding a convenient approach for \emph{fully automated}
reasoning.

@@

\noindent 
All modal connectives are simply lifted versions of the
usual logical connectives. Notations are used to allow the modal
connectives to be used as similarly as possible to the usual
connectives. The prefix ``\texttt{m}'' is used to distinguish the
modal connectives: if $\odot$ is a connective on type \text{Prop},
$\texttt{m}\odot$ is a connective on the lifted type $\texttt{o}$ of
modal propositions.

@@

%\noindent
To maximally preserve user intuition in interactive modal logic
theorem proving, the embedding via the possible world semantics should
be as transparent as possible to the user. Fortunately, the basic Coq
tactics described above automatically unfold the shallowest modal
definition in the goal.  Therefore, they can be used with modal
connectives and quantifiers just as they are used with the usual
connectives and quantifiers. The situation for the new modal
operators, on the other hand, is not as simple, unfortunately.

@@

\noindent If the new modal tactics above are regarded from a natural
deduction point of view, they correspond to the inference
rules shown in Figure
\ref{fig:NDK}. Because of this correspondence and the Henkin-completeness of the modal natural deduction calculus\footnote{
  The ND calculus with the rules from
  Figures \ref{fig:NaturalDeduction} and \ref{fig:NDK} is sound and
  complete relatively to the calculus of Figure
  \ref{fig:NaturalDeduction} extended with a necessitation rule and the
  modal axiom K \cite{ModalND}. Starting from a sound and 
  Henkin-complete ND calculus for classical higher-order logic 
  (cf. Figure~\ref{fig:NaturalDeduction}), the additional
  modal rules in Figure \ref{fig:NDK} make it sound and Henkin-complete
  for the rigid higher-order modal logic \textbf{K}.}, 
the tactics allow the user to prove any valid modal formula without having 
to unfold the definitions of the modal operators.

The labels that name boxes in the inference rules of Figure \ref{fig:NDK} are
precisely the worlds that annotate goals and hypotheses in \Coq with
the modal embedding. A hypothesis of the form \texttt{(p w)}, where
\texttt{p} is a modal proposition of type \texttt{o} and \texttt{w} is
a world of type \texttt{i} indicates that \texttt{p} is an assumption
created inside a box with name \texttt{w}.

@@

\s\s
\infer[\nec_I]{\nec A}{\omega: \fbox{\infer*{A}{}} }
\s
\infer[\nec_E]{w: \fbox{ \infer*{}{A} } }{\nec A}
%\vspace{2em}
\s
\infer[\pos_E]{\omega: \fbox{ \infer*{}{A} } }{\pos A}
\s
\infer[\pos_I]{\pos A}{w: \fbox{\infer*{A}{}} }

@@

In order to demonstrate the efficacy and convenience of the modal
embedding approach not only for proving simple lemmas and theorems,
but also for larger developments, we include here a full and detailed
formalization of G\"odel's ontological argument. which has been verified in Coq 8.4pl5.

@@

%The theological implications of the verified correctness of G\"odel's
% proof certainly depend on a critical discussion of the underlying
% concepts, definitions and axioms, which is beyond the scope of this
% paper.  Clearly, the application of theorem proving technology to the
% domain of theoretical philosophy can --- as already pictured by
% Leibniz --- be very fruitful for both areas.
